---
title: "Assessing Factors that Affect mpg"
author: "PilotGtec EFTL"
date: "21 January 2015"
output: html_document
---

Analyze mtcars data. The dataset has 32 observations on 11 variables, namely `mpg` (miles per gallon), `cyl`, `disp`, `hp`, `drat`, `wt`, `qsec`, `vs`, `am`, `gear`, and `carb`. Details on the variable descriptions can be found [here][1]. In particular, the automobile industry is interested in the following:

- “Is an automatic or manual transmission better for MPG”
- "Quantify the MPG difference between automatic and manual transmissions"

[1]: https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html "here" 

#Executive Summary

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
data(mtcars)
mtcars$cyl <- factor(mtcars$cyl)
mtcars$am <- factor(mtcars$am)
mtcars$vs <- factor(mtcars$vs)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
model3a <- lm(mpg ~ ., data = mtcars)
model3 <- step(model3a, direction = "both")
summ3 <- summary(model3)
data(mtcars)
```

We explore the `mtcars` data and determine the effects of various factors to the `mpg` values of automobiles, particularly focusing on the type of transmission (as an initial approach). Three regression models are set up and compared (using anova). 

- Model 1: Transmission type only
- Model 2: Transmission and weight
- Model 3: Stepwise algorithm

Results show that **Model 3** (`lm(formula = mpg ~ cyl + disp + wt + am, data = mtcars)`) is the most optimal. **Model 3** captures `r summ3$r.squared*100`% of the the total variance  (`r summ3$adj.r.squared*100`%, adjusted). From the coefficient values, a `1000-lb` increase in an autombile's weight would result to a `r summ3$coefficients[5]` change in the `mpg`. With respect to the transmission, automatic transmission has `r summ3$coefficients[6]` less mpg than manual. For `hp`, on the other hand, an increase would result to a `r summ3$coefficients[4]` (decrease) in `mpg`. Finally, in general, more cylinders (`cyl`) would decrease the automobile's `mpg`. Analysis of residuals is also presented in this report.

##Exploratory Data Analysis
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
correlations <- (cor(mtcars$mpg, select(mtcars, cyl:carb)[,]))
```

First we look at the different relationships of each of the pair of variables in the dataset by plotting the pairs (Figure A1, *Appendix*). We also mark those `x~y` pairs that fit (`r^2 > 0.700`) into  a linear model. In the plot, we have excluded *cyl*, *vs*, *am*, *gear*, and *carb* as they are categorical variables--- we plot them independently (Figure A2, *Appendix*). The plots give us an idea about the correlations of the variables (*Note: categorical variables are considered numerical*). To quantify the correlations, we implement the `cor` function with `wt` (`r correlations[5]`), `cyl` (`r correlations[1]`), `disp` (`r correlations[2]`), and `hp` (`r correlations[3]`) showing the highest correlation values. 


## Question 1: “Is an automatic or manual transmission better for MPG”
```{r, echo=FALSE, results='hide', message=FALSE}
library(plyr)
ttest<-t.test(mtcars$mpg~mtcars$am,conf.level=0.95)
means <- ddply(mtcars,~am,summarise,mean=mean(mpg),sd=sd(mpg))
mtcars$cyl <- factor(mtcars$cyl)
mtcars$am <- factor(mtcars$am)
mtcars$vs <- factor(mtcars$vs)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
```

From Figure A2, we see that there is a relationship between MPG and the type of transmission an automobile has: **Automatic** has lower *mpg* than **Manual** with mean values `r means$mean[1]` and `r means$mean[2]`, respectively. To further **quantify the statistical significance of the difference**, we conduct a *`r ttest$method`* on the categories with a $95\%$ confidence interval. This results to a `p-value` of `r ttest$p.value` suggesting that indeed **Automatic** has lower *mpg* than **Manual**. We have to note, however, that there are other variables that need to be considered as well. This is where the regression models come in. 

##The Models
```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
all <- lm(mpg~., data = mtcars)
step <- step(all, direction= "both")
an<-anova(all, step)
```

*Note: Categorical variables are converted from numerical to factor.*

In this section, we explore three models listed below. Here, we excluded the model (**Model 0**) involving all variables. This was decided after implementing the `anova` function comparing it against the model (**Model 3**) that resulted from a stepwise algorithm performed on **Model 0**. Even though **Model 3** is not significantly better (`Pr(>F) ~` `r an[6][2,]`) than **Model 0**, it has fewer variables.

- Model 1: Transmission type only
- Model 2: Transmission and weight
- Model 3: Stepwise algorithm

###Comparing the Models

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
model1 <- lm(mpg ~ am, data = mtcars)
model2 <- lm(mpg ~ wt + am, data = mtcars)
model3a <- lm(mpg ~ ., data = mtcars)
model3 <- step(model3a, direction = "both")
c1<-summary(model1)$coefficient
r1<-summary(model1)$r.squared
c2<-summary(model2)$coefficient
r2<-summary(model2)$r.squared
c3<-summary(model3)$coefficient
r3<-summary(model3)$r.squared
summ3 <- summary(model3)
```

Computing an anova for the three models shows that **Model 3** is the most optimal model compared to **Model 2** and **Model 1**. 

```{r, echo=FALSE}
an<-anova(model1, model2, model3)
print(an)
```

**Model 3** captures `r summ3$r.squared*100`% of the the total variance and `r summ3$adj.r.squared*100`% (adjusted). From the coefficients, we can see that a `1000-lb` increase in an autombile's weight would result to a `r summ3$coefficients[5]` change to the `mpg`. With respect to the transmission, automatic transmission has `r summ3$coefficients[6]` less mpg than manual. In terms of `hp`, an increase would  result to a `r summ3$coefficients[4]` (decrease) in `mpg`. Finally, in general, more cylinders (`cyl`) would decrease the automobile's `mpg`.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summ3$coefficients
```

In the Appendix section (Figure A3), we analyze and plot the residuals against the fitted and leverage values for **Model 3**. We also provide the Normal Q-Q plot and Scale-Location plots. 


##Appendix

###Figure A1

```{r, fig.height=5, warning=FALSE, message=FALSE, echo=FALSE}
library(dplyr)
lmfit = function(x,y) {
    points(x,y,pch="o")
    pVal <- anova(lm(y~x))$'Pr(>F)'[1]
    rSquared <- summary(lm(y~x))$r.squared
    if ( rSquared > 0.6500){
        abline(lm(y~x),col="red", lwd=2)
    }
    }
pairs(select(select(mtcars, mpg:qsec),-cyl), panel = lmfit)
```

###Figure A2

```{r, echo=FALSE, warning=FALSE, fig.height=6}
par(mfrow=c(3,2))
with(mtcars, {
    plot(cyl, mpg, main="Cylinders", xlab="", ylab="mpg")
})
with(mtcars, {
    plot(am, mpg, main="Transmission", xlab="", ylab="mpg")
})
with(mtcars, {
    plot(vs, mpg, main="V/S", xlab="", ylab="mpg")
})
with(mtcars, {
    plot(gear, mpg, main="Forward Gears", xlab="", ylab="mpg")
})
with(mtcars, {
    plot(carb, mpg, main="Carburetors", xlab="", ylab="mpg")
})
```

###Figure A3

```{r echo=FALSE, warning=FALSE, message=FALSE}
plotgrid <- par(mfrow = c(2, 2))
plot(model3)
par(plotgrid)
```

